Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-09 ms.tex
2.2

\section{Model}
\label{sec:model}
\subsection{Galaxy models}
\label{sec:gal_model}
\citet{LauerFaber+:2007a} use Hubble Space Telescope WFPC2 imaging to
measure the radial surface brightness profiles for hundreds of nearby
early type galaxies. The measured brightness profile is generally
well-fit by a ``Nuker" law parameterization:
\begin{equation}
  I(\xi)=I_b 2^{(\beta-\Gamma)/\alpha} \xi^{-\Gamma} (1+\xi^\alpha)^{-(\beta-\Gamma)/\alpha}, \,\,\,\xi\equiv\frac{r}{r_b},
\end{equation}
i.e., a broken power law that transitions from an inner power law
slope, $\Gamma$, to an outer power law slope, $\beta$, at a
break radius, $\rb$.  Assuming the stellar population is spherically
symmetric, this corresponds to a stellar density $\rhostar \propto
r^{-1-\Gamma}$ for $r \ll \rb$ and $\rhostar\propto r^{-1-\beta}$ for
$r \gg \rb$.  We can write the deprojected stellar density approximately (formally, this is the $\alpha \rightarrow \infty$ limit) as
\begin{align}
\rho_\star = 
\begin{cases}
\rho_\star|_{\rsoi} \left(r/\rsoi\right)^{-1-\Gamma} & r \leq r_b\\
\rho_\star|_{r_b} \left(r/r_b\right)^{-1-\beta} & r > r_b,
\label{eq:rhostar}
\end{cases}
\end{align}
where $\rho_\star|_{\rsoi}$ is the stellar density at the radius of the black hole sphere of influence\footnote{This approximate expression agrees surprisingly well with the mean empirical scaling relation for $r_{\rm inf}(M_\bullet)$ calculated in Appendix C of \citet{Stone&Metzger15}, although we note that this relation has significant scatter.}, 
\begin{equation}
\rsoi \simeq G \Mbh/\sigma^2 \approx 14 M_{\bullet,8}^{0.6}\,{\rm pc},
\label{eq:rsoi}
\end{equation}
where $M_{\bullet,8} \equiv M_{\bullet}/10^{8}M_{\odot}$ and the
second equality in (\ref{eq:rsoi}) employs the $\Mbh-\sigma$
relationship of \citet{McConnellMa+:2011a},
 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}
Though of questionable validity for low mass black holes (e.g., \citealt{Greene&Ho07}), we adopt equation (\ref{eq:Msigma}) for lack of an obviously better alternative.

@@

We fix $\beta = 2$ because our results are
generally not sensitive to the properties of the gas flow on radial
scales $\gtrsim r_b$. The presence of the break radius $r_{b}$ is,
however, necessary to obtain a converged steady state for some regions
of our parameter space.  We consider solutions calculated for up to
four values of $\rb$: 50 pc, 100 pc, 200 pc, and 400 pc, motivated by
the range of break radii from the \citet{LauerFaber+:2007a}
sample.\footnote{For the core galaxies the break radius follows the
  scaling relationship $\rb\sim 106 \, (\Mbheight)^{0.39}$ pc, with
  scatter of approximately one dex.  Most the cusp galaxies have $\rb$
  between 100 and 1000 pc, and lack a clear trend with $\Mbh$.}  In
practice we calculate solutions for a range of $\rb$ only in cases
when its precise value is expected to have a significant affect on the
solution, in particular for high SMBH masses and low heating rates,
for which the stagnation radius approaches large radii. We also
neglect value of $r_b$ which would give unphysically large bulges for
a given $\Mbh$.

@@

The source term in equation (\ref{eq:drhodt}),
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 1.4 \times 10^{10}$ yr.
To good approximation $\eta\simeq 0.02 (\tau_{\star}/t_h)^{-1.3}$ at
time $\tau_{\star}$ following an impulsive star burst (e.g.,
\citealt{Ciotti+91})\footnote{\citet{Ciotti+91} give the mass return
  rate from evolved stars as a function of B-band luminosity instead
  of volumetrically, but our expressions are equivalent.}, although
$\eta$ is significantly higher for continuous SFHss
(bottom panel of Fig.~\ref{fig:vwSources}).

Source terms $\propto q$ also appear in the momentum and entropy
equations (eqs.~[\ref{eq:dvdt}] and [\ref{eq:dsdt}]) because the isotropic injection of mass represents, in the SMBH rest frame, a source of momentum
and energy relative to the mean flow.  Physically, these result from
the mismatch between the properties of virialized gas injected locally
by stellar winds and the mean properties of the background flow.  The
term $\propto -q c_{s}^{2}$ is important because it acts to potentially stabilize the flow against runaway cooling ($\S\ref{sec:instability}$).

The term $\propto \vw^2 = \sigma(r)^2+v_{w}^2$ in the entropy equation
accounts for external heating sources (e.g.,
\citealt{ShcherbakovWong+:2014a}).  The first term approximates the
minimum amount of shock heating from stellar winds due to the random
movement of stars with respect to one another at the characteristic
velocity of 
\be\sigma \approx \sqrt{\frac{3 G \Mbh}{(\Gamma+2)
    r}+\sigma_0^2},
\label{eq:sigmarel}
\ee where the first term is the velocity dispersion of the power law
stellar density profile in the SMBH potential.  The second term,
$\sigma_0\approx 3 \sigma^2$ is a constant, motivated by the observed
velocity dispersion outside of the SMBH sphere of influence
\citep{ChaeBernardi+:2014a}, where $\sigma$ is the velocity dispersion
in the $\Mbh-\sigma$ relation (eq.~[\ref{eq:Msigma}]).  The second
component of the total heating, $v_{w}^{2}$, parameterizes all
additional sources of energy input, including faster winds from young
stars, millisecond pulsars, supernovae, AGN feedback, etc.  We assume
that $v_w$ is constant with radius, i.e. that the volumetric heating
rate is proportional to the local stellar density.  The value of $v_w$
will in general depend on the SMBH mass and the age of the stellar
population ($\S\ref{sec:heating}$).

>>>>

\section{Model}
\label{sec:model}
\subsection{Galaxy models}
\label{sec:gal_model}
\citet{LauerFaber+:2007a} use Hubble Space Telescope WFPC2 imaging to
measure the radial surface brightness profiles for hundreds of nearby
early type galaxies. The measured profile is fit by a ``Nuker" law parameterization:
\begin{equation}
  I(\xi)=I_b 2^{(\beta-\Gamma)/\alpha} \xi^{-\Gamma} (1+\xi^\alpha)^{-(\beta-\Gamma)/\alpha}, \,\,\,\xi\equiv\frac{r}{r_b},
\end{equation}
i.e., a broken power law that transitions from an inner power law
slope, $\Gamma$, to an outer power law slope, $\beta$, at a
break radius, $\rb$.  If the stellar population is spherically
symmetric, then this corresponds to a 3D stellar density $\rhostar \propto
r^{-1-\Gamma}$ for $r \ll \rb$ and $\rhostar\propto r^{-1-\beta}$ for
$r \gg \rb$.  We can write the deprojected stellar density approximately (formally, this is the $\alpha \rightarrow \infty$ limit) as
\begin{align}
\rho_\star = 
\begin{cases}
\rho_\star|_{\rsoi} \left(r/\rsoi\right)^{-1-\Gamma} & r \leq r_b\\
\rho_\star|_{r_b} \left(r/r_b\right)^{-1-\beta} & r > r_b,
\label{eq:rhostar}
\end{cases}
\end{align}
where $\rho_\star|_{\rsoi}$ is the stellar density at the radius of the black hole sphere of influence\footnote{This approximate expression agrees surprisingly well with the mean empirical scaling relation for $r_{\rm inf}(M_\bullet)$ calculated in Appendix C of \citet{Stone&Metzger15}, although we note that this relation has significant scatter.}, 
\begin{equation}
\rsoi \simeq G \Mbh/\sigma^2 \approx 14 M_{\bullet,8}^{0.6}\,{\rm pc},
\label{eq:rsoi}
\end{equation}
where $M_{\bullet,8} \equiv M_{\bullet}/10^{8}M_{\odot}$ and the
second equality in (\ref{eq:rsoi}) employs the $\Mbh-\sigma$
relationship of \citet{McConnellMa+:2011a},
 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}
Though of questionable validity for low mass black holes (e.g., \citealt{Greene&Ho07}), we adopt equation (\ref{eq:Msigma}) for lack of an obviously better alternative.

@@

We fix $\beta = 2$ because our results are not overly sensitive to the properties of the gas flow on radial scales $\gtrsim r_b$. The presence of the break radius $r_{b}$ is,
however, necessary to obtain a converged steady state for some regions
of our parameter space.  We consider solutions calculated for up to
four values of $\rb$: 50 pc, 100 pc, 200 pc, and 400 pc, motivated by
the range of break radii from the \citet{LauerFaber+:2007a}
sample.\footnote{For the core galaxies the break radius follows the
  scaling relationship $\rb\sim 106 \, (\Mbheight)^{0.39}$ pc, with
  scatter of approximately one dex.  Most the cusp galaxies have $\rb$
  between 100 and 1000 pc, and lack a clear trend with $\Mbh$.}  In
practice we calculate solutions for a range of $\rb$ only in cases
when its precise value is expected to have a significant affect on the
solution, in particular for high SMBH masses and low heating rates,
for which the stagnation radius approaches large radii. We also
neglect value of $r_b$ which would give unphysically large bulges for
a given $\Mbh$.

@@

The source term in equation (\ref{eq:drhodt}),
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 1.4 \times 10^{10}$ yr.
To good approximation $\eta\simeq 0.02 (\tau_{\star}/t_h)^{-1.3}$ at
time $\tau_{\star}$ following an impulsive star burst (e.g.,
\citealt{Ciotti+91})\footnote{\citet{Ciotti+91} give the mass return
  rate from evolved stars as a function of B-band luminosity instead
  of volumetrically, but our expressions are equivalent.}, although
$\eta$ is significantly higher for continuous SFHs
(bottom panel of Fig.~\ref{fig:vwSources}).

Source terms $\propto q$ also appear in the momentum and entropy
equations (eqs.~[\ref{eq:dvdt}] and [\ref{eq:dsdt}]) because the isotropic injection of mass represents, in the SMBH rest frame, a source of momentum
and energy relative to the mean flow.  Physically, these result from
the mismatch between the properties of virialized gas injected by stellar winds and the mean background flow.  The
term $\propto p/\rho = c_{s}^{2}$ is important because it acts to stabilize the flow against runaway cooling ($\S\ref{sec:instability}$).

The term $\propto \vw^2 = \sigma(r)^2+v_{w}^2$ in the entropy equation
accounts for external heating sources (e.g.,
\citealt{ShcherbakovWong+:2014a}), where
\be\sigma \approx \sqrt{\frac{3 G \Mbh}{(\Gamma+2)
    r}+\sigma_0^2},
\label{eq:sigmarel}
\ee 
is the stellar velocity dispersion and which accounts for the minimal amount of shock heating from stellar winds due to the random motion of stars in the SMBH potential, while $\sigma_0^{2} \approx 3 \sigma^2$ is assumed to be constant, corresponding to the observed
velocity dispersion outside of the SMBH sphere of influence
\citep{ChaeBernardi+:2014a}.  The second term, $v_{w}^{2}$, parameterizes additional sources of energy input, including faster winds from young
stars, millisecond pulsars, supernovae, AGN feedback, etc  ($\S\ref{sec:heating}$).  We assume
that $v_w$ is constant with radius, i.e. that the volumetric heating
rate is proportional to the local stellar density.  
%The value of $v_w$
%will in general depend on the SMBH mass and the age of the stellar
%population ($\S\ref{sec:heating}$).
